Synaptic connectome of the ctenophore apical organ

### 2025-02-13 manuscript/Manuscript.qmd
balancer_plot_3d
MS: add hypothes.is

---
title: "Cell-type diversity and connectome of the ctenophore apical organ"
mainfont: Arial
monofont: DejaVu Sans Mono #Times New Roman
fontsize: 11pt
linestretch: 1.2
linkcolor: blue
documentclass: article
max-width: 100%
format:
  html: default
  pdf:
    pdf-engine: xelatex
    number-sections: false
    margin-left: 0.8in
    margin-right: 0.8in
    margin-top: 0.8in
    margin-bottom: 0.8in
    toc: false
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{lineno}
    include-before-body:
      text: |
        \linenumbers
bibliography: references.bib #add bibliography in this file
csl: elife.csl #change according to journal
---

\

Kei Jokura^1,3,\*^, Sanja Jasek^1,2^, Lara Kewalow^2^, Pawel Burkhardt^4^, Gáspár Jékely^1,2,\*^\

^1^Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom\
^2^Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany\
^3^National Institute for Basic Biology, Okazaki, ..., Japan
^4^Sars Centre ... Berge, Norway
^\*^Correspondence: corr_author\@email.com

@@

First descriptions of AO: [@aronova1974electron]

The lobate ctenophore *Mnemiopsis*

The lithocytes have a large calcium carbonate concretion and... [@Tamm_2014]
Living lithocytes are transported along the balancer cilia to the tip of the statocyst [@Noda_2014]
The projections of a group of bridge cells located on two sides of the apical organ forms an arch and connects the balancers along the tentacular plane [@Tamm_2002].

The easiest way is to use the command line to add new refs to the .bib file:

```{bash}
#| echo: true
#| eval: false

#curl -LH "Accept: application/x-bibtex" https://doi.org/10.7554/eLife.91258.1 >> references.bib
```

@@

- specimen 
- volume, resolution, fixation, quality
- number of sections
- skeletonisation in Catmaid
- number of cells
- general anatomy, regions, quadrants

@@

### Classification of cell types

- ultrastructural criteria (cilia, vesicles etc)
- number of types, cells per type - table

```{r, out.width = "100%", fig.cap="**Figure 2. Title fig 2** Balancer complex: balancer, lithocyte, groove 2 types, bridge - EM pics, (A) legend (B) legend. " }
#| echo: false

@@

"the larval bridge has not yet developed forks at its ends, as seen in adult Mnemiopsis bridges"[@Tamm_2002]

Figure3 - dome, bicil, monocil, no-cil, plumose, dense vesicle cells (rename), gland? cells, oligo/multiciliated cells

Plumose cells [@Hernandez_Nicaise_1984] may be pressure sensors [@Tamm_2014b]

Figure4 - lamellate body complex, LB, intra-multi-ciliated, relationship to balancer

Figure5 - nerve net, different fragments/cells/syntytia, EM, relation to entire organ

Figure 6 -  cilia? renderings of basal bodies, centrioles, length, axoneme structure (table), table about cell types, number of cilia etc.

## Synapses

@@

@@

## Sourcing code and working with variable

```{r}
#| include: false
#| eval: true
#source(here::here("analysis/scripts/statistics_for_paper.R"))
```

The 'analysis/scripts/statistics_for_paper.R' script is sourced and it runs but the output is not included in the knitted output. But we can access the variables defined in the sourced script simply by adding \` r var_name \` between \` backticks, in this case max_PRC value is (now this number comes from our sourced script).

If we update the data, the script can recalculate the variable we want to refer to in the text and update the number.

@@

@@

We would like to thank the Jekely lab for the R project template (<https://github.com/JekelyLab/new_paper_template>) we used to write this paper. This work was funded by ...
JSPS...
ERC..
Others ...

## References

>>>>

---
title: "Synaptic connectome of the ctenophore apical organ"
subtitle: |
  \raggedright  Kei Jokura^1,2,4,5\*^, Sanja Jasek^1,3,4^,  Lara Kewalow^3^, Pawel Burkhardt^6^, Gáspár Jékely^1,3,4,\*^  \newline<br>^1^Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom \newline<br>^2^Exploratory Research Center on Life and Living Systems (ExCELLS), National Institutes of Natural Sciences, Okazaki, Japan\newline<br>^3^Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany \newline<br>^4^BioSciences, University of Exeter, Stocker Road, Exeter EX4 4QD, Exeter, United Kingdom\newline<br> ^5^National Institute for Basic Biology (NIBB), National Institutes of Natural Sciences, Okazaki, Japan\newline<br>^6^Michael Sars Centre, University of Bergen, Norway \newline<br> ^\*^Correspondence:  jokura@nibb.ac.jp, gaspar.jekely@cos.uni-heidelberg.de
mainfont: Arial
monofont: DejaVu Sans Mono #Times New Roman
fontsize: 11pt
linestretch: 1.2
linkcolor: blue
documentclass: article
max-width: 100%
format:
  html:
    embed-resources: true 
  pdf:
    pdf-engine: xelatex
    number-sections: false
    margin-left: 0.8in
    margin-right: 0.8in
    margin-top: 0.8in
    margin-bottom: 0.8in
    toc: false
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{lineno}
    include-before-body:
      text: |
        \linenumbers
bibliography: references.bib #add bibliography in this file
csl: elife.csl #change according to journal
---

```{r}
#| include: false
#| eval: true
#source(here::here("analysis/scripts/statistics_for_paper.R"))

#The 'analysis/scripts/statistics_for_paper.R' script is sourced and it runs but the output is not included in the knitted output. But we can access the variables defined in the sourced script simply by adding \` r var_name \` between \` backticks, in this case max_PRC value is (now this number comes from our sourced script).

```

@@

Understanding how neurons evolved and what their initial functions were is key to elucidating the processes by which animals developed flexible behaviors. Neurons do not function in isolation; rather, they form networks that integrate and process information, enabling the emergence of complex behaviors. Therefore, understanding the structure and function of neural circuits is essential for investigating the mechanisms underlying animal behavior.

Recent comparative genomic studies (Ryan et al., 2013; Whelan et al., 2015; Whelan et al., 2017; Li et al., 2021; Whelan and Halanych, 2023) and chromosome-level synteny analyses across different animal phyla (Schultz et al., 2023) suggest that ctenophores are the sister group to all other metazoans and were the first animals to evolve a nervous system. This has led to attempts to compare genes and molecular modules in ctenophores using single-cell analyses, in which neuropeptides have been employed as neuronal markers (Sebé-Pedrós et al., 2018; Sachkova et al., 2021; Hayakawa et al., 2022). However, the homologous relationships between ctenophore genes and those of other animals are highly divergent, limiting the feasibility of sequence-based comparisons. Given these challenges, we considered an alternative approach: linking neural circuits to physiological responses to gain insight into neural function.

Despite their phylogenetic position, ctenophores exhibit sophisticated and dynamic behavioral patterns in response to a wide range of external stimuli. These behaviors are thought to be regulated by multiple neural systems, including the subepithelial nerve net (SNN), mesogleal neurons, an apical organ, and tentacle nerves. However, the mechanisms by which these neural systems control behavior remain largely unexplored.

Among these, we focused on the apical organ of ctenophores. In many bilaterian and non-bilaterian animals, an apical nervous system is located at one pole of the body, forming a specialized sensory structure that governs physiological functions and motor activities. In ctenophores, ultrastructural analyses have revealed that the aboral pole contains a diverse array of sensory cells and neuronal components (Hernandez-Nicaise, 1968, 1973, 1974; Tamm, 1982). Additionally, immunostaining using an anti-tyrosylated α-tubulin antibody has identified a neural structure called the deep nerve net within the apical organ, suggesting its involvement in neural regulation of sensory-motor control (Jager et al., 2011).

In this study, we specifically examined the gravity-sensing system within the apical organ. The apical organ of ctenophores contains a dense mineralized mass known as the statocyst, supported by four mechanosensory ciliated cells called balancer cells [@Noda_2014]. The ciliary movements of these balancer cells function as pacemakers for the eight rows of comb plates that control posture and swimming through mechanotransduction [@Tamm_2014] (Tamm, 1982). Synaptic connections projecting to the balancer cells have also been reported [@aronova1974electron]. Therefore, we conducted a connectome analysis to characterize the network structure of the apical organ and elucidate the neural circuits that link sensory input to behavioral responses in ctenophores.

@@

First, we embedded the entire body of the *Mnemiopsis leidyi* cydippid-stage larvae, five days post-fertilization, in Epon resin following high-pressure freezing. To reconstruct all synaptic-resolution connections among the cells comprising the apical organ, also known as the aboral organ, we prepared approximately 1,000 ultra-thin serial sections (50 nm thick) from the aboral tip of the larval body embedded in the resin block. Using a scanning electron microscope (Gemini SEM 500) with a resolution of 2.0 nm/pixel, we imaged only the region containing the apical organ. Subsequently, 620 of these images were stitched and aligned using TrakEM2. The resulting volumetric EM dataset had dimensions of 60 μm × 40 μm × 30 μm. We traced and annotated all cells in this dataset, ultimately reconstructing 909 cells, each with both nuclei and cell bodies intact.

Cells containing cilia were traced down to the tips of the cilia, and basal bodies were annotated accordingly. For neuronal skeletonization, nodes were placed to interconnect the profiles of the same neuron’s processes across layers, extending the skeleton until all branches were fully traced. Each node was tagged, and skeletons were named and assigned multi-level annotations. As described later, many neurons we identified formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main trunk or other branches. In such cases, branch nodes were placed near the closest existing node and annotated accordingly. The entire skeletonized volume was composed of 134,591 nodes. However, 88 fragments could not be attached to somata-associated skeletons. Most of these fragments represented short skeletal branches that could not be traced beyond gaps or low-quality layers.

Next, we decided to divide the entire apical organ into four broad quadrants to facilitate grouping the identified cells. The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the positions of the anal pores. This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development. We categorized the traced cells into four quadrants and designated these groups as Q1 through Q4 for clarity and consistency.

@@

### Identification of Synaptic Structures in Syncytium Neurons

Classical neural staining techniques do not provide clear images of the neurons at the aboral pole. However, ultrastructural studies have provided morphological evidence of elements resembling neurons, based on synaptic structures, located on the epithelial floor of the apical organs [@Horridge_1964; @Hernandez_Nicaise_1973]. Based on these findings, we identified synaptic structures characteristic of ctenophore neurons in our data, using previously identified pre-synaptic triad morphological features, such as single-layered vesicles, smooth endoplasmic reticulum, and tightly packed mitochondria. In our study, we specifically identified synaptic sites and marked mitochondria (orange) as synaptic nodes. The regions where synaptic vesicles align were marked as connectors (light blue arrows) between cells across the membrane. These connectors link the synaptic nodes of the pre-synaptic cytoskeleton to the partner nodes of the post-synaptic cytoskeleton. As previously reported, the specialization of post-synaptic structures in ctenophores is not apparent, so we recognized synaptic vesicle clusters on the pre-synaptic membrane as the point of reference. Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells.

As mentioned above, following the identification of the presynaptic structures, we reconstructed three major Syncytial neurons. Each of these Syncytial neurons was a cell with multiple nuclei, with membranes fused by continuous plasma membranes. These neurons are distinct in morphology from the Syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously reconstructed in 3D [@Burkhardt_2023]. Our findings represent the second documented discovery of Syncytial neurons in ctenophores. Furthermore, these neurons exhibited clear morphological differences when compared to other sensory cells reported in the same study, such as mesogleal neurons and sensory cells with presynaptic structures and cilia (types 1-4)[@Burkhardt_2023]. Based on their distinct spatial relationships, we were able to classify these three Syncytial neurons into two categories. The first type is a larger "AO neuron_Q1234," which possesses four (or possibly six?) nuclei spanning four quadrants. It contained X presynaptic structures. The second type consists of "AO neuron_Q12" and "AO neuron_Q34," each having two nuclei spanning two quadrants, and each containing X presynaptic structures.

```{r, out.width = "100%", fig.cap="**Figure 2. Title fig 2** pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions, (A) legend (B) legend. " }
#| echo: false

@@

### Identification the Gravity-Sensitive Neural Circuit via the Syncytial Neurons Network

We discovered that AO neurons form synaptic connections with each other, and that AO neuron_Q1234 forms self-synapses, or “autapses.” To our knowledge, previous reports have not identified synaptic connections between subepithelial nerve net (SNN) neurons. Thus, our results represent the first report of synaptic connectivity between neurons in the apical organ, forming a network in ctenophores (subject to confirmation). These AO neuron networks were found to form many presynaptic structures in relation to the gravity-sensing balancer cells.

Balancer cells are monociliated cells, and their cilia protrude longer than those of other monociliated cells. Moreover, the cilia of balancer cells are bundled into four groups at the center of the apical organ, forming a compound cilium. Based on these features, we identified and classified the balancer cells. The cellular arrangement differed clearly when viewed in the lateral view, sagittal plane, and tentacular plane, with the cell bodies gathering toward the apical organ in the tentacular plane. In each quadrant, the number of cells was as follows: Q1: 37, Q2: 32, Q3: 32, Q4: 28. Each cell contained 3 to 10 ? mitochondria.

From AO neuron_Q1Q2, synapses were formed with 6 of the 37 balancer cells in the Q1 region, and 8 of the 32 balancer cells in the Q2 region. Similarly, from AO neuron_Q3Q4, synapses were formed with 1 of the 32 balancer cells in the Q3 region, and 5 of the 28 balancer cells in the Q4 region. AO neuron_Q1234 formed input synapses with balancer cells in the Q1 region (7 cells), Q2 region (11 cells), Q3 region (6 cells), and Q4 region (10 cells). Some balancer cells received inputs from both AO neuron_Q1Q2 or AO neuron_Q3Q4 and AO neuron_Q1234. While previous studies have suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez_Nicaise_1974], our data did not reveal any synaptic inputs from balancer cells to AO neurons. However, we identified "bridge cells" that actively formed input synapses with AO neurons, highlighting their potential role in the network.

```{r, out.width = "100%", fig.cap="**Figure 3. Title fig 3** pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions, (A) legend (B) legend. " }
#| echo: false

knitr::include_graphics("figures/Figure3.png")
```

### Identification of Neuron-like Bridge Cells Forming Afferent Synapses in the Gravity-Sensitive Neural Circuit

From our data, we identified several cell groups that form multiple afferent synapses with AO neurons. These cells, based on their morphological features, were found to be the bridge cells first described by Tamm et al. in 2002 [@Tamm_and_Tamm_2002]. These cells are characterized by bundles of elongated processes filled with microtubules that arch over the epithelial layer, resembling a bridge. They originate from the base of paired balancer cells along the tentacle surface and extend across the sagittal plane toward the base of the opposite balancer cells. In regions where the mitochondria of the bridge cells are localized (approximately 30%), a presynaptic triad structure, similar to that of AO neurons, was found, containing synaptic vesicles and smooth endoplasmic reticulum.

Our three-dimensional reconstruction data revealed that the bridge cells form two distinct cell groups across the sagittal plane, between the Q1Q2 and Q3Q4 regions. In the Q1Q2 region, 14 cells were identified, while in the Q3Q4 region, 12 cells were identified, totaling 26 bridge cells. Nearly all of these bridge cells (25 out of 26) exhibited afferent synapses from AO neurons. For bridge cells located in the Q1Q2 region, synaptic inputs came primarily from AO neuron_Q1Q2 (11 cells), AO neuron_Q1234 (1 cell), or both (2 cells). For bridge cells located in the Q3Q4 region, synaptic inputs were received from AO neuron_Q3Q4 (1 cell), AO neuron_Q1234 (7 cells), or both (1 cell). In other words, bridge cells in the Q1Q2 region mainly received input from AO neuron_Q1Q2, while those in the Q3Q4 region received input primarily from AO neuron_Q1234, showing a distinct pattern of input in the two regions.

Some bridge cells also formed afferent synapses with AO neurons. For example, bridge cells in the Q1Q2 region formed synapses with AO neuron_Q1Q2 (3 cells) or both AO neuron_Q1Q2 and AO neuron_Q1234 (2 cells). Bridge cells in the Q3Q4 region formed synapses with AO neuron_Q3Q4 (1 cell) or AO neuron_Q1234 (6 cells). A notable difference was observed between the Q1Q2 and Q3Q4 regions in the proportion of synaptic inputs from bridge cells to AO neurons.

Interestingly, in both regions, bridge cells also formed synapses with adjacent bridge cells. However, no synaptic input was found from bridge cells across the sagittal plane to those in the opposite region. To analyze the grouped synaptic connectivity graph, we classified cell types within each region, collapsed cells of the same type into a single node, summed the number of synapses, and explored directed pathways to effectors (balancer cell groups). The results revealed a feedback pathway from AO neurons through the synaptic connections formed by bridge cells, thereby shedding light on the neural circuit structure involving these cells.

```{r, out.width = "100%", fig.cap="**Figure 4. Title fig 4** pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions, (A) legend (B) legend. " }
#| echo: false

knitr::include_graphics("figures/Figure4.png")
```

### Control of Balancer Ciliary Movements by Gravity-Sensitive Neural Circuits

To investigate the function of gravity-sensitive circuits, we conducted a comparative analysis of balancer ciliary movements across different regions. Previous studies (Tamm, 1980, 1982; Lowe, 1997) have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. Furthermore, it has been suggested that differences in the asymmetry of statolith morphology and the elongation direction of balancer cilia between the tentacular and sagittal planes could result in variations in the force received from the statolith (Tamm, 2014, 2015). Building on these findings, we standardized conditions for observing balancer cilia by tilting the microscope by 90 degrees and fixing samples such that the aboral-oral axis was aligned parallel to the vertical stage, with the aboral side facing upward. Using connectome analysis, we aimed to evaluate the neural circuit differences by analyzing the control patterns of left and right balancer cilia in the tentacular and sagittal planes. Specifically, we hypothesized that differences in control between Q1 (Q2) and Q3 (Q4) regions, as well as between Q1 (Q4) and Q2 (Q3) regions, would reveal distinct functions of AO neurons (e.g., AO neuron Q12/Q34 vs. AO neuron Q1234). For the sagittal plane, we recorded balancer ciliary movements in the Q1(4) and Q2(3) regions (left and right sides of the image) and, for the tentacular plane, in the Q1(3) and Q2(4) regions, using a high-speed camera (100 fps) for 2 minutes (12,000 frames). Regions of interest (ROIs) were defined in areas where brightness changes indicated ciliary beating. We measured these brightness changes to generate dynamics graphs of balancer ciliary movements for both planes.

During the 2-minute recordings, balancer cilia exhibited movements that could be fast, slow, or stop abruptly (arrest) and start moving again (re-beat). Occasionally, large contraction responses caused the entire cydippid to move out of frame, and data from these moments were excluded. We focused on two metrics for left-right comparison: (1) the timing of ciliary arrest and (2) the timing of re-beat initiation following an arrest. Due to the low baseline beating frequency of the cilia, rapid synaptic input responses were not evaluated, suggesting that ciliary beat regulation may involve non-synaptic control mechanisms. The analysis revealed that both arrest and re-beat events were synchronized between left and right balancers in both planes within a few milliseconds, suggesting that these events are governed by synaptic input from neurons. Interestingly, while the re-beat timing showed no significant differences between the sagittal and tentacular planes, arrest timing exhibited greater variability in the tentacular plane. Comparing these findings with neural circuit diagrams suggests that in the sagittal plane, the same AO neurons (Q12 or Q34) project to both balancer cell groups, transmitting signals simultaneously and enabling synchronization. In contrast, in the tentacular plane, separate AO neurons project to different regions, potentially introducing an additional synaptic step, which may account for the observed variability in arrest timing.

```{r, out.width = "100%", fig.cap="**Figure 5. Title fig 5** pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions, (A) legend (B) legend. " }
#| echo: false

knitr::include_graphics("figures/Figure5.png")
```

## Discussion

In this study, we traced all the constituent cells of the aboral organ in five-day-old larvae of *Mnemiopsis leidyi* and completed a comprehensive neural map. As a result, we identified a novel type of syncytial neuron with a previously unreported morphology. Furthermore, we found that three such syncytial neurons project to distinct quadrants, integrating with balancer cell groups to form a gravity-sensing neural circuit. This study represents the first complete neural circuit map reported in ctenophores. Additionally, our analysis of balancer ciliary movements, based on neural circuit data, revealed differences in ciliary arrest control arising from circuit variations.

### Identification of a Previously Unreported Syncytial Neuron

We identified three syncytial neurons with multiple nuclei in the apical organ of five-day-old *M. leidyi* cydippid larvae. Burkhurdt et al. (2023) previously reported that in one-day-old larvae, the sub-epithelial nerve net (SNN) contains neurons that form syncytia. However, the newly discovered syncytial neurons embedded within the apical organ cell cluster did not exhibit the characteristic "pearl necklace-like structure" of the SNN, which consists of microtubules connecting neurons. Instead, their morphology was more reminiscent of fiber cells previously described in Placozoa (Mayorova et al., 2021), with cell bodies extending through intercellular spaces. These morphological characteristics suggest that the newly identified neurons represent a second distinct type of syncytial neuron in ctenophores.

### Functional Diversification of the Three Syncytial Neurons

The three identified neurons may be classified into two distinct functional types based on their spatial arrangement and their influence on the differential control of balancer ciliary movement between the quadrants. In metazoans, a single neuron can directly provide synaptic input to multiple ciliated cells at a distance, synchronizing their movements. For instance, in larvae of the annelid *Platynereis dumerilii*, a cholinergic motor neuron (MC) innervates multiple prototroch ciliated cells, synchronizing their motion such that activation of the neuron leads to simultaneous arrest of ciliary beating (Verasztó et al., 2017). Additionally, a pair of serotonergic neurons, designated Ser-h1, extend axons that cross at the midline; the left Ser-h1_l predominantly innervates right-sided ciliated cells, while the right Ser-h1_r mainly projects to the left-side ciliated cells. Activation of Ser-h1 induces ciliary beating initiation and increases beat frequency.

In the this study, we found that three neurons are closely positioned within the confined space of the apical organ, potentially enabling a single neuron to control distant ciliary movements in a synchronized manner. Moreover, the presence of neurons projecting to the same ciliated cells in distinct patterns suggests that these neurons may serve opposing functions—one inhibiting ciliary activity while the other promotes it, or one modulating ciliary beat frequency. While our synaptic ultrastructure analysis did not allow us to distinguish excitatory from inhibitory neurons, future electrophysiological investigations will be essential to elucidate their functional properties.

```{r, out.width = "100%", fig.cap="**Figure 6. Title fig 6** pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions, (A) legend (B) legend. " }
#| echo: false

knitr::include_graphics("figures/Figure6.png")
```

### Paracrine Signaling and Synaptic Transmission from Sensory to Neural Cells

Previous electron microscopy studies have suggested the presence of various sensory cells, including photoreceptor and mechanoreceptor cells, within the aboral organ of ctenophores (Horridge, 1965; Krisch, 1973; Aronova, 1974; Vinnikov, 1974). By comparing our dataset with previously reported ultrastructural data, we successfully identified several sensory cells within the apical organ. However, notably, we did not find any clear presynaptic structures between these sensory cells and either neural or other cellular targets. In prior studies, immunostaining for specific neuropeptides in ctenophores has revealed positive signals in distinct cell populations within the apical organ (Sachkova et al., 2021; Hayakawa et al., 2022). These findings suggest that many sensory cells in the apical organ may primarily rely on neuropeptide-mediated paracrine signaling rather than direct synaptic connections, supporting the chemical brain hypothesis (Jekely, 2021). While paracrine transmission is generally slower than synaptic transmission, it allows for a greater diversity of information processing.

Conversely, previous research has demonstrated that mechanoreceptive sensory cells equipped with cilia (referred to as type 3 and type 4 cells) in ctenophores form direct synaptic connections with the sub-epithelial nerve net (SNN) and mesogleal neurons (Burkhurdt et al., 2023). Based on their ultrastructural features, these sensory cells are believed to respond to physical stimuli such as water flow, vibrations, and direct contact, transmitting information rapidly to cilia and muscle cells (Hertwig, 1880; Chun, 1880; Eimer, 1880).

Surprisingly, our study identified bridge cells as the primary cell type forming synaptic inputs onto neurons in the apical organ. Bridge cells were first described by Tamm et al. (2002) based on their morphological characteristics, with cellular projections extending to the basal region of balancer cilia [@Tamm_2002] (Tamm & Tamm, 2002). Our findings suggest that bridge cells are electrically coupled to balancer cilia and provide rapid feedback to neurons via synaptic transmission. Future studies analyzing differential responses to various sensory stimuli will not only elucidate the functional roles of bridge cells but also provide insights into the evolutionary relationship between transmission speed and synapse development.

### Circuit Variability as a Driver of Behavioral Diversity

Our study suggests that differences in neural circuits contribute to variations in the timing of ciliary arrest, which may play a role in modulating swimming direction in ctenophores (Richard Satterlie, 2015). This finding provides evidence that increased circuit complexity contributes to behavioral diversification. By integrating multiple neurons and altering circuit structure, information can be branched, allowing for diverse motor patterns within the framework of fast synaptic transmission.

The neural cells identified in the apical organ exhibit a divergent feedforward connectivity pattern, where a single neuron forms presynaptic connections with multiple downstream targets. Such a circuit design minimizes unwanted signal interference, reduces processing delays, and enables relatively fast response execution (Luo, 2021). Moreover, the presence of multiple divergent feedforward circuits suggests that slight variations in the synchronization and phase delay of ciliary beating (within the range of a few hundred milliseconds) may have evolved as a mechanism for generating behavioral diversity.

From a broader neuroanatomical perspective, nervous system structures across various animals have been shaped through selective optimization to adapt to environmental constraints (Bullmore & Sporns, 2012; Perin et al., 2011). For example, in nematodes, the spatial arrangement of ganglia is optimized to minimize the total length of neural wiring (White et al., 1986). Similarly, in ctenophores, understanding how transmission via syncytial neurons and synaptic connections has been selectively optimized at the circuit level will provide crucial insights into the evolutionary scaling of neural networks from early nervous systems to complex brains (Farnworth and Montgomery, 2024). In particular, this research may help elucidate how circuit modularization and layered structures emerged over evolutionary time.

### Data and Code Availability

The following EM image stacks, including all traces and annotations, are available at https://catmaid.jekelylab.ex.ac.uk. These datasets can be queried either within CATMAID or through the CATMAID application programming interface (API) using the catmaid or natverse packages (Bates et al., 2020). The dataset encompasses all EM images (in JPG format), skeletons, meshes, node tags, connectors, and annotations. Additionally, we provide all R scripts used for data acquisition and figure generation (Jokura et al., 2025). All plots, figures (including anatomical renderings), and figure layouts can be fully reproduced using the provided R scripts. While the scripts are mostly organized by figure, some general-purpose scripts—for tasks such as loading libraries, accessing CATMAID data, and defining common functions—are shared across multiple figures.

### Specimen Preparation, Transmission Electron Microscopy, and Image Processing

Larvae of Mnemiopsis leidyi (5 days old) were cryofixed using a high-pressure freezing apparatus (BAL-TEC HPM 010, Balzers) and immediately transferred to liquid nitrogen for storage. The frozen samples were processed in a substitution medium containing 2% (w/v) osmium tetroxide and 0.5% uranyl acetate in acetone, using a cryo-substitution device (EM AFS-2, Leica). Cryo-substitution was performed by gradually raising the temperature, and the samples were embedded in epoxy resin. Serial sections of 50 nm thickness were prepared using a Reichert Jung Ultracut E ultramicrotome and a 45º DiATOME diamond knife. To enhance section adhesion and improve hydrophilicity, a conductive indium tin oxide-coated glass slide (ITO Glass, UQG Optics) was treated with air glow discharge using the PELCO easiGlow system (Ted Pella, Inc.), rendering the carbon film surface negatively charged. Section ribbons were collected on the prepared glass slides, slowly dried to ensure proper stretching, and firmly adhered to the glass surface. The sections on glass were stained with UranyLess and lead citrate (Reynolds) using airless staining bottles (Delta Microscopies). The glass slides were mounted on STEM-specific stages (Zeiss) using Copper Foil EMI Shielding Tape (3M). Imaging was performed using a Gemini SEM 500 (Zeiss) equipped with SmartSEM and Atlas 5 imaging software (Zeiss). The full dataset consisted of 620 serial sections, each 50 nm thick. The imaging resolution was 2.8 nm/pixel, using electron holography transmission at an acceleration voltage of 1.5 kV (in-lens detector, dwell time: 3 µs).

### Image-Stack Realignment and Spatial Data Transformation in CATMAID

To process the image stack, we utilized the TrakEM2 plugin of FIJI (ImageJ) (version 2.0.0-rc-15/1.49k / Java 1.6.0_24 (64-bit) – 2014). A project was created, and all TIFF images were imported using the "import sequence as grid" function. Subsequently, the following filters were applied sequentially: Invert, Equalize Histogram, and Gaussian Blur.

The alignment process consisted of three stages, each progressively refining the spatial accuracy:

Rigid Alignment Initially, a rigid alignment was performed with the following parameters: least squares mode (linear feature correspondences), encompassing the entire layer range with the first layer as the reference. Only visible images were used, without propagation. The alignment was executed with an initial Gaussian blur of 1.6 pixels, three steps per scale octave, a minimum image size of 512 pixels, and a maximum of 2048 pixels. Additional parameters included a feature descriptor size of 8, orientation bins of 8, and a closest ratio of 0.92. The alignment allowed clearing the cache, using 32 feature extraction threads, a maximal alignment error of 100 pixels, a minimal inlier ratio of 0.20, and a minimum of 12 inliers. The expected and desired transformations were set to rigid, with testing multiple hypotheses (tolerance: 5.00 pixels) and considering up to 5 neighboring layers, giving up after 5 failures. Regularization was performed with a maximal iteration of 1000, a maximal plateau width of 200, and a rigid lambda of 0.10.

Affine Alignment Following the rigid alignment, an affine alignment was applied using similar parameters, except the expected and desired transformations were set to affine. The minimal image size was reduced to 64 pixels, while the other parameters (e.g., Gaussian blur, feature descriptor size, inliers, and testing hypotheses) remained unchanged to ensure consistent processing.

Elastic Alignment Two iterations of elastic alignment were performed to fine-tune the spatial data. Key parameters included a block matching layer scale of 0.05, a search radius of 200 pixels, a block radius of 2000 pixels, and a resolution of 60. Correlation filters were employed with a minimal PMCC r of 0.10, a maximal curvature ratio of 1000, and a maximal second-best r/best r of 0.90. A local smoothness filter was applied with the approximate local transformation set to affine, a local region sigma of 1000 pixels, and an absolute maximal local displacement of 10 pixels (relative maximal displacement: 3.00). Pre-aligned layers were tested for up to 4 neighboring layers. The elastic alignment used a rigid approximation, maximal iterations of 3000, a plateau width of 200, spring mesh stiffness of 0.01, and a maximal stretch of 2000 pixels. A legacy optimizer was employed to enhance performance.

After each alignment stage, the project was saved as an XML file under a unique name to preserve iterative progress. Finally, the images were exported from FIJI using TrakEM2 in a format compatible with CATMAID.

### Neuron Tracing, Synapse Annotation, and Review

The reconstruction of cells and connectivity was performed using CATMAID (Saalfeld et al., 2009), analyzing a total of 620 serial sections. To digitally reconstruct all neurons within the serial TEM dataset, we utilized the collaborative web-based application CATMAID, installed on a local server (Saalfeld et al., 2009; Schneider-Mizell et al., 2016). To mark the locations of cell bodies, we placed tags at the approximate center of each nucleus within the dataset. At each nuclear center, the radius of the single node was adjusted according to the size of the cell body in that specific layer. All skeletons were rooted at their respective cell bodies, and nodes were tagged as "soma." Synapses were identified based on four key structural features: the cell membrane, synaptic vesicles, the endoplasmic reticulum, and mitochondria. Most synapses could be verified across consecutive sections, ensuring accurate annotation and connectivity mapping.

### Cell Nomenclature and Annotations

Each cell is assigned a specific type name based on its category (balancer, biC, bridge, bristle, dense vesicle cell, dome, EF, IcMC, LB, lithocyte, monoC, non-C, plumose cell, SNN, ssCG, SSN, stCG), resulting in a total of 17 cell types. Additionally, a quadrant number is appended to the cell type name following an underscore ("\_") to indicate the cell body’s location. For example, cells located in the first quadrant are annotated with "Q1." If a cell is situated between the first and second quadrants, the annotation "Q1Q2" is applied.

Cells of the same type within the same region are further distinguished by serial numbering. Each cell is also assigned multiple annotations, which can be utilized to query the database via CATMAID or the CATMAID API (e.g., using the R catmaid package). These annotations provide a structured and precise framework for identifying and analyzing specific cells, facilitating robust data integration and retrieval within the dataset.

This system ensures consistent and accurate tracking of cellular data, enabling comprehensive analysis of cellular structures and functions across regions.

### Analysis of Balancer Cilia

We conducted the analysis using the cydippid stage of Mnemiopsis leidyi at five days post-fertilization. A coverslip with a thin layer of Vaseline applied to two edges was lightly placed on a slide glass. Filtered natural seawater was introduced under the coverslip, along with the cydippids. By gently moving the coverslip, the desired orientation for observation was adjusted, and once positioned, slight pressure was applied to immobilize the cydippids. To stabilize the statolith's position, the microscope was tilted 90 degrees, arranging the stage vertically. The movement of the balancer cilia was observed using a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a highly sensitive CMOS monochrome camera optimized for the near-infrared (NIR) range (UI-3360CP-NIR-GL Rev.2, iDS). Observations were made using a 40x objective lens (Objective LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27) with glycerine immersion. For recording, we employed a custom-made NIR LED strobe illumination system (wavelength: 850 nm) synchronized with the camera's exposure signals. The camera and LED strobe system were operated using the Video Capture Software BohNavi. Images were recorded at a resolution of 640×480 pixels with a frame rate of 100 fps for 2 minutes, using 0.05 ms pulses from the LED. To analyze the ciliary beating of the balancer cilia, we utilized the Multi Kymograph project in Fiji software.
=======
Figure 6 - cilia? renderings of basal bodies, centrioles, length, axoneme structure (table), table about cell types, number of cilia etc.

### Synapses

@@

### Connectivity

### Physiology?

### Discussion

@@

@@

### Volume electron microscopy

### Imaging of balancer activity

### Cell-type annotation and connectome analysis

### Data and code availability

>>>>>>> Stashed changes

@@

This work was funded by ... JSPS... This project has received funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement No 101020792).

## References {.unnumbered}

::: {#refs}
:::

```{=html}
<!--
## See https://github.com/hypothesis/client
## Add commenting functionality with this script
-->
```
```{=html}
<script src="hypothes.is/embed.js" async></script>
```
\newpage

## Supplementary material {.unnumbered}
